---
title: "Regression Models Course Project"
author: "Brian Hampton"
date: '2022-06-11'
output:
  pdf_document: default
  html_document: default
---
## Executive Summary

This report addresses questions posed in the Johns Hopkins Coursera Regression Models course project:

*Working for Moter Trend, a magazine about the automobile industry, analyze the collection*
*of cars in the mtcars data set to answer:*


*1. Is an automatic or manual transmission better for MPG?*

*2. Quantify the MPG difference between automatic and manual transmissions.*

Manual transmission cars have a better MPG rating.  A univariate model, which only 
accounts for 36 percent of the observed variability, predicts manual transmission cars
to be better than automatic transmission cars by 7.2 gallons.  A multivariate model,
which accounts for 85% of the observed variability predicts manual transmission cars to
be better than automatic transmission cars by 2.9 mpg. Further investigation into the 
selection of cars within the mtcars data set may reveal additional factors that contributed
to the quantified results.

## Exploratory Data Analysis
```{r setup, echo=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(knitr)
library(datasets)
library(dplyr)
data("mtcars")
```
*NOTE: code not shown in the report body is included in the appendix*

After looking at the mtcars data set, for easier interpretation I added a variable
*Transmission* with responses of *Automatic* or *Manual* based on the *am* variable
responses.

```{r change am to Transmission, echo=FALSE}
mtcars$Transmission_Type[mtcars$am==0] <- "Automatic"
mtcars$Transmission_Type[mtcars$am==1] <- "Manual"
df <- mtcars[,-9, drop=FALSE]
```
**Table 1. Summary of Miles per Gallon (mpg) by Transmission Type**
```{r Table 1, echo=FALSE}
table_1 <- df %>%
    group_by(Transmission_Type) %>%
    summarise(n = n(),
              min = min(mpg),
              q1 = quantile(mpg, 0.25),
              median = median(mpg),
              mean_mpg = mean(mpg),
              q3 = quantile(mpg, 0.75),
              max = max(mpg),
              sd_mpg = sd(mpg))
kable(table_1, digits = 2)
```

```{r box plot 1, echo=FALSE}
bp1 <- ggplot(df, aes(x=Transmission_Type, y=mpg, fill=Transmission_Type, 
                          group=Transmission_Type)) + geom_boxplot()
bp1 <- bp1 + geom_dotplot(binaxis='y', stackdir='center', dotsize=1)
bp1 <- bp1 + labs(x="Transmission Type", y="Miles per Gallon (mpg)", 
title="Miles per Gallon (mpg) by Transmission Type")
bp1
```

The summary table for mpg by transmission and the visual representation of that data suggests
that manual transmission cars within the data set have a higher mpg.  

## Inference
### Null Hypothesis
H~o~: there is no difference in mean mpg between automatic and manual transmission cars.

H~a~: the mean mpg for manual transmission cars is greater than the mean mpg for automatic transmission cars.

### Univariate Regression Model
```{r univariate regression, echo=TRUE}
uv <- lm(mpg ~ Transmission_Type, data = df)
summary(uv)
```
Considering only the relationship between mpg and transmission type, manual transmission cars 
are more fuel efficient, by about 7.2 mpg. However, the R-squared value is 0.3598, meaning
that only 36 percent of the observed variability is explained by this univariate regression model.

### Multivariate Regression Model
```{r multivariate regression, echo=TRUE}
mv <- step(lm(mpg ~ ., data=df),direction="both", trace=FALSE)
summary(mv)
```
```{r multivariate result for average car, echo=TRUE}
mv_a <-lm(mpg ~ I(wt - mean(wt)) + I(qsec - mean(qsec)) + Transmission_Type, data = df)
summary(mv_a)$coef
```

The multivariate model indicates that weight (wt), and quarter-mile time (qt) are confounding
variables to the transmission type and mpg relationship within this data set.  With an R-squared
value 0f 0.8497, 85% of the observed variability in the data set is explained by this 
multivariate model.  With this model a manual transmission car is expected to have an mpg rating
of 2.9 higher/better than an automatic transmission car. For a mean weight car with a mean
quarter-mile time, the model expects a manual transmission car to have an mpg of 21.8 and an
automatic transmission car to have an mpg of 18.9. 

### Residual Analysis

```{r residual plot, echo=FALSE}
ra <- ggplot(mv, aes(.fitted, .resid)) + geom_point()
ra <- ra + stat_smooth(method="loess", col="red") + geom_hline(yintercept=0, linetype="dashed")
ra <- ra + xlab("Fitted Values") + ylab("Residuals")
ra <- ra + ggtitle("Residual vs Fitted Plot") + theme_bw()
ra
```
The residuals appear to "bounce randomly" around the 0 line, suggesting that it is reasonable
to assume a linear relationship.  The residuals roughly form a "horizontal band" around the
0 line except for the very highest observed mpg observations, suggesting that the majority
of the error terms are equal.  No one residual appears to stand out from the pattern of
residuals, suggesting that there are no outliers.

### Uncertainty in Results
Given report length constraints I recommend further investigation to better understand the
data set. Specifically why are 7 of the 13 manual transmission cars lighter than the lightest
manual transmission car. And conversely, why are 8 of the 19 automatic transmission cars
heavier than the heaviest manual transmission car. When looking at a subset of mtcars bounded 
by the lightest automatic transmission car and the heaviest manual transmission car, the difference
in mpg by transmission type is not as discernible.

**Table 2. Summary of MPG by Transmission Type for Data Bounded by Lightest Automatic and Heaviest Manual Transmission Cars**
```{r table of mtcars bounded by lightest auto & heaviest manual, echo=FALSE}
auto <- subset(df, Transmission_Type == "Automatic")
manual <- subset(df, Transmission_Type == "Manual")
df2 <- subset(df, wt >= min(auto$wt))
df2 <- subset(df2, wt <= max(manual$wt))
table_2 <- df2 %>%
    group_by(Transmission_Type) %>%
    summarise(n = n(),
              min = min(mpg),
              q1 = quantile(mpg, 0.25),
              median = median(mpg),
              mean_mpg = mean(mpg),
              q3 = quantile(mpg, 0.75),
              max = max(mpg),
              sd_mpg = sd(mpg))
kable(table_2, digits = 2)
```

### Conclusion
1. Manual transmission cars attained a higher mpg within the entire mtcars data set.
2. A model that considers weight and quarter-mile time as confounding variables over the entire mtcars
data set accounts for more of the observed variability within the data set.

Other factors not included in the variables, such as manufacturing trends or people's
preferences, may have played a role in the selection of the mtcars data set and influenced
the results.

## Appendix
```{r code for mtcars data disply}
knitr::opts_chunk$set(echo = TRUE, results = "hide")
### Initial look at mtcars
display <- head(mtcars, 2)
```

```{r code to change variable names}
mtcars$Transmission_Type[mtcars$am==0] <- "Automatic"
mtcars$Transmission_Type[mtcars$am==1] <- "Manual"
df <- mtcars[,-9, drop=FALSE]
```

```{r code for Table 1, echo=TRUE}
### Summary table 1 for mtcars mpg by Transmission
table_1 <- df %>%
    group_by(Transmission_Type) %>%
    summarise(n = n(),
              min = min(mpg),
              q1 = quantile(mpg, 0.25),
              median = median(mpg),
              mean_mpg = mean(mpg),
              q3 = quantile(mpg, 0.75),
              max = max(mpg),
              sd_mpg = sd(mpg))
kable(table_1, digits = 2)
```

```{r code for box plot 1, echo=TRUE}
### box plot 1 for mtcars mpg by Transmission
bp1 <- ggplot(df, aes(x=Transmission_Type, y=mpg, fill=Transmission_Type, 
                          group=Transmission_Type)) + geom_boxplot()
bp1 <- bp1 + geom_dotplot(binaxis='y', stackdir='center', dotsize=1)
bp1 <- bp1 + labs(x="Transmission Type", y="Miles per Gallon (mpg)", 
title="Miles per Gallon (mpg) by Transmission Type")
```

```{r code for residual plot, echo=TRUE}
### Residual vs Fitted Plot for the multivariate (mv) model
ra <- ggplot(mv, aes(.fitted, .resid)) + geom_point()
ra <- ra + stat_smooth(method="loess", col="red") + geom_hline(yintercept=0, linetype="dashed")
ra <- ra + xlab("Fitted Values") + ylab("Residuals")
ra <- ra + ggtitle("Residual vs Fitted Plot") + theme_bw()
```

```{r code fortable of mtcars bounded by lightest auto & heaviest manual, echo=TRUE}
### Summary table 2: subset mtcars bounded by lightest automatic and heaviest manual
### and publish summary table of mpg by transmission type
auto <- subset(df, Transmission_Type == "Automatic")
manual <- subset(df, Transmission_Type == "Manual")
df2 <- subset(df, wt >= min(auto$wt))
df2 <- subset(df2, wt <= max(manual$wt))
table_2 <- df2 %>%
    group_by(Transmission_Type) %>%
    summarise(n = n(),
              min = min(mpg),
              q1 = quantile(mpg, 0.25),
              median = median(mpg),
              mean_mpg = mean(mpg),
              q3 = quantile(mpg, 0.75),
              max = max(mpg),
              sd_mpg = sd(mpg))
kable(table_2, digits = 2)